Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-11-30 paper/paper_no_event.tex
Checked numbers in paper. Minor edits.

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. 
%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. 
%With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can accurately measure quantities such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of NS radii and the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

@@

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the effective one body (EOB) formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. In Ref.~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, direct comparisons between this code and traditional samplers have not been published for BNS systems using the same waveform models. It is therefore not clear how accurate this code is at measuring quantities such as the tidal parameters. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

@@

We organize the paper as follows. In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section~\ref{sec:surrogate}, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}. In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.

@@

In this Section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions. We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$. This is favorable for data analysis which is usually done in the frequency domain. It also allows us to extend the model down to arbitrarily low frequencies with an analytic, frequency-domain model. Unfortunately, Fourier transforming a finite length waveform leads to a surrogate with more noise. We will show below, however, that sufficient filtering can solve this problem. The \texttt{SEOBNRv4T} waveform model, as implemented, works for mass ratios in the range $q \in [1/3, 1]$, spins $\chi_{1,2} \in [-0.5, 0.5]$ and tidal parameters $\Lambda_{1,2} \in [0, 5000]$. Our surrogate of this model, \texttt{SEOBNRv4T\_surrogate}, will be valid for the same range of parameters.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At a starting frequency of 20~Hz, over an hour is needed to evaluate a single \texttt{SEOBNRv4T} waveform on a standard CPU, so this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is Gaussian process regression (GPR)~\cite{RasmussenWilliams2006}.

@@

Given the GPR fits for the amplitude and phase residuals, we can now reconstruct the frequency-domain waveform. We label the set of GPR fits at the frequency nodes $MF_j$ by $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ for the amplitude residuals and by $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ for the phase residuals. The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
\begin{align}
\Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}](Mf).
\end{align}
We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to \texttt{TaylorF2} at lower frequencies. With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
\begin{align}
\label{eq:amp_sur}
A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right], \\
\label{eq:phase_sur}
\Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
\end{align}
In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are 
\begin{align}
\label{eq:hplus_sur}
\tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
\label{eq:hcross_sur}
\tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right].
\end{align}

@@

Fig.~\ref{fig:uncsamp} shows the estimated RMS error $\epsilon_{\rm RMS}(\bx)$ maximized over the parameters $\bx$ for each new sample added to the training set. This is done for both the Mat\'{e}rn (black) and squared exponential (gray) kernels. For each kernel, we constructed an initial surrogate with optimized hyperparameters, then fixed the hyperparameters before choosing the 400 new samples using uncertainty sampling. The non-smoothness of the curves results because the basin-hopping algorithm does not always find the global maximum of $\epsilon_{\rm RMS}$ at each iteration. We find that increasing the number of waveforms from 160 to 560 decreases the global maximum of $\epsilon_{\rm RMS}$ by a factor of \red{$\sim 4$} for the Mat\'{e}rn kernel and a factor of \red{$\sim 10$} for the squared exponential kernel. In Section~\ref{sec:accuracy} we will examine how well this estimated error agrees with the true error.

@@

The accuracy of the surrogate can be assessed by comparing it to a test set of waveforms. We generate 1000 waveforms with the original parameters sampled uniformly in the ranges $q \in [1/3, 1]$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$. In Fig.~\ref{fig:uncsamp}, we show box plots of the RMS phase error of the test set waveforms for each of 8 surrogate models generated. The first two surrogates are constructed using the initial design with either the Mat\'{e}rn or squared exponential kernel. The second two surrogates are constructed with the initial design and the 400 uniformly distributed samples and either the Mat\'{e}rn or squared exponential kernel. The next two surrogates use the initial design and the 400 samples chosen by the uncertainty sampling algorithm for each kernel. The final two surrogates use all 960 waveforms for each kernel. We find that for the Mat\'{e}rn kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ is a reasonable estimate of the maximum test-set RMS phase errors. However, for the squared exponential kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ significantly underestimates the true maximum test-set error. Additionally, we find that the surrogates constructed with the uniform distribution have smaller median test-set errors. However, the surrogates constructed with uncertainty sampling have smaller maximum test-set errors.

@@

In Fig.~\ref{fig:mismatch}, we show box plots of the mismatches $\mathcal{M}$ between the surrogate and the 1000 test-set waveforms for each of our 8 surrogates. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and a total mass of $M=2.8M_\odot$. Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of \red{15.2}~Hz and \red{5076}~Hz respectively. As we increase the number of samples in our training set, the mismatches decrease. However, it is less clear which kernel and sampling strategy is optimal. For our final surrogate, we simply choose the surrogate that has the smallest maximum and median mismatches, and this is the surrogate with the Mat\'{e}rn kernel and all 960 waveforms. This surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$}.

@@

In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for a equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor. The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz. At large starting frequencies the evaluation time is limited by the \red{$\sim 0.01$~s} needed to calculate the GPR fits at each frequency node. As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed. The evaluation time is then dominated by the spline interpolation needed to resample the waveform. We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817 (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} and the BBH model \texttt{SEOBNRv4\_ROM}. Only the analytic \texttt{TaylorF2} model is noticeably faster. In the bottom panel, we compare the evaluation time to the original time-domain \texttt{SEOBNRv4T} waveform model and find it is $\sim 1500$--6000 times faster.

@@

%\subsection{Injection runs}

As a final test of the surrogate, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template. We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}. We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the EOS named MS1b in~\cite{ReadLackey2009}. The sky location and inclination angle are fixed at an arbitrary value, and the distances are chosen such that the optimal network signal-to-noise ratios (SNRs) are either 30, 60, or 120. We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.

>>>>

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

@@

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

In this paper, we will use an implementation of the EOB formalism that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note there is an alternative EOB model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a different point-particle and spin prescription. For matter effects, this model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances.

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. In Ref.~\cite{BNSPE}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems. It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions. The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. In this work, instead of using a reduced basis, we will instead use the fact that the Fourier-transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

@@

We organize the paper as follows. In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section~\ref{sec:surrogate} we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}. In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms. Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.

@@

In this section we describe how we decompose the EOB waveform into smooth, slowly-varying functions and train a surrogate model for these functions. We will work with the frequency-domain waveform $\tilde h(Mf; \bx)$. This is favorable for data analysis which is usually done in the frequency domain. It also allows us to extend the model down to arbitrarily low frequencies with an analytic, frequency-domain model. Unfortunately, Fourier transforming a finite length waveform leads to a surrogate with more noise. We will show below, however, that sufficient filtering can solve this problem. The \texttt{SEOBNRv4T} waveform model, as implemented, works for mass ratios in the range $q \in [1/3, 1]$, spins $\chi_{1,2} \in [-0.5, 0.5]$ and tidal parameters $\Lambda_{1,2} \in [0, 5000]$. Our surrogate of this model, \texttt{SEOBNRv4T\_surrogate}, will be valid for the same range of parameters.

@@

Next we choose a method to interpolate the residuals $\Delta\ln(A)(MF_j; \bx)$ and $\Delta\Phi(MF_j; \bx)$ at each of the frequency nodes $MF_j$ as a function of the five waveform parameters $\bx$. Most multivariate interpolation techniques (e.g. tensor spline or Chebyshev interpolation) require a function to be sampled on a rectangular grid, and thus suffer from the curse of dimensionality: the number of samples grows exponentially with the dimension $d$ ($N^d$ samples for $N$ points per dimension). For our 5-dimensional problem, $10^5$ waveform evaluations are needed for only 10 samples per parameter. At a starting frequency of 20~Hz, over an hour is needed to evaluate a single \texttt{SEOBNRv4T} waveform on a standard CPU, so this is at the limit of what is reasonable. However, there are other methods that do not require a rectangular grid, and this allows us to choose more efficient experimental designs as discussed in Sec.~\ref{sec:design} below. The method we choose is GPR~\cite{RasmussenWilliams2006}.

@@

Given the GPR fits for the amplitude and phase residuals, we can now reconstruct the frequency-domain waveform. We label the set of GPR fits at the frequency nodes $MF_j$ by $\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}$ for the amplitude residuals and by $\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}$ for the phase residuals. The surrogates for the residuals $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$ are constructed by interpolating between the nodes $MF_j$ with cubic splines:
\begin{align}
\Delta\ln A_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\ln A_j](\bx)\}](Mf),\\
\Delta\Phi_S(Mf;\bx) &= \mathcal{I}_{\rm Spline}[\{\mathcal{I}_{\rm GPR}[\Delta\Phi_j](\bx)\}](Mf).
\end{align}
We set these functions to 0 below the first frequency node $MF_0 =  Mf_{{\rm trunc},i}$ so that the waveform transitions to TaylorF2 at lower frequencies. With the analytic expressions for $A_{\rm F2}(Mf;\bx)$ and $\Phi_{\rm F2}(Mf;\bx)$ and the interpolated expressions $\Delta\ln A_S(Mf;\bx)$ and $\Delta\Phi_S(Mf;\bx)$, the final surrogates for the amplitude and phase are
\begin{align}
\label{eq:amp_sur}
A_S(Mf;\bx) &= A_{\rm F2}(Mf;\bx) \exp\left[ \Delta\ln A_S(Mf;\bx) \right], \\
\label{eq:phase_sur}
\Phi_S(Mf;\bx) &= \Phi_{\rm F2}(Mf;\bx) + \Delta\Phi_S(Mf;\bx).
\end{align}
In physical units, for an inclination angle $\iota$, the $+$ and $\times$ polarizations of the waveform are 
\begin{align}
\label{eq:hplus_sur}
\tilde h_+(f; \bx) &= \frac{1}{2}(1+\cos^2\iota) \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right)\right], \\
\label{eq:hcross_sur}
\tilde h_\times(f; \bx) &= \cos\iota \frac{G^2 M^2}{c^5 d} A_S\left(\frac{GMf}{c^3}; \bx\right) \nonumber \\
& \times \exp\left[i \Phi_S\left(\frac{GMf}{c^3}; \bx\right) + i \frac{\pi}{2}\right].
\end{align}

@@

Fig.~\ref{fig:uncsamp} shows the estimated RMS error $\epsilon_{\rm RMS}(\bx)$ maximized over the parameters $\bx$ for each new sample added to the training set. This is done for both the Mat\'{e}rn (black) and squared exponential (gray) kernels. For each kernel, we constructed an initial surrogate with optimized hyperparameters, then fixed the hyperparameters before choosing the 400 new samples using uncertainty sampling. The non-smoothness of the curves results because the basin-hopping algorithm does not always find the global maximum of $\epsilon_{\rm RMS}$ at each iteration. We find that increasing the number of waveforms from 160 to 560 decreases the global maximum of $\epsilon_{\rm RMS}$ by a factor of $\sim 4$ for the Mat\'{e}rn kernel and a factor of $\sim 10$ for the squared exponential kernel. In Section~\ref{sec:accuracy} we will examine how well this estimated error agrees with the true error.

@@

The accuracy of the surrogate can be assessed by comparing it to a test set of waveforms. We generate 1000 waveforms with the original parameters sampled uniformly in the ranges $q \in [1/3, 1]$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$. In Fig.~\ref{fig:uncsamp}, we show box plots of the RMS phase error of the test set waveforms for each of the 8 surrogate models. The first two surrogates are constructed using the initial design with either the Mat\'{e}rn or squared exponential kernel. The second two surrogates are constructed with the initial design and the 400 uniformly distributed samples and either the Mat\'{e}rn or squared exponential kernel. The next two surrogates use the initial design and the 400 samples chosen by the uncertainty sampling algorithm for each kernel. The final two surrogates use all 960 waveforms for each kernel. We find that for the Mat\'{e}rn kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ is a reasonable estimate of the maximum test-set RMS phase errors. However, for the squared exponential kernel, the global maximum of $\epsilon_{\rm RMS}(\bx)$ significantly underestimates the true maximum test-set error. Additionally, we find that the surrogates constructed with the uniform distribution have smaller median test-set errors. However, the surrogates constructed with uncertainty sampling have smaller maximum test-set errors.

@@

In Fig.~\ref{fig:mismatch}, we show box plots of the mismatches $\mathcal{M}$ between the surrogate and the 1000 test-set waveforms for each of our 8 surrogates. We use the design sensitivity aLIGO PSD~\cite{Aasi:2013wya} and a total mass of $M=2.8M_\odot$. Our integration bounds are $Mf_{\rm low} = 0.00021$ and $Mf_{\rm high} = 0.07$, corresponding to physical units of 15.2~Hz and 5076~Hz respectively. As we increase the number of samples in our training set, the mismatches decrease. However, it is less clear which kernel and sampling strategy is optimal. For our final surrogate, we simply choose the surrogate that has the smallest maximum and median mismatches, and this is the surrogate with the Mat\'{e}rn kernel and all 960 waveforms. This surrogate has a maximum mismatch of $4.5\times 10^{-4}$.

@@

In Fig.~\ref{fig:timing}, we show the surrogate evaluation time as a function of the starting frequency $f_{\rm min}$ for an equal mass $1.4M_\odot$--$1.4M_\odot$ binary evaluated on a 3.5~GHz Intel Xeon processor. The waveform was sampled with a uniform frequency spacing corresponding to a sampling rate of 4096~Hz. At large starting frequencies the evaluation time is limited by the $\sim 0.01$~s needed to calculate the GPR fits at each frequency node. As $f_{\rm min}$ decreases, the waveform becomes longer and a smaller frequency spacing is needed. The evaluation time is then dominated by the spline interpolation needed to resample the waveform. We also compare the evaluation time to three BNS waveform models used recently in the analysis of GW170817 (\texttt{TaylorF2}, \texttt{SEOBNRv4\_ROM\_NRTidal}, \texttt{IMRPhenomD\_NRTidal})~\cite{BNSPE} and the BBH model \texttt{SEOBNRv4\_ROM}. Only the analytic \texttt{TaylorF2} model is noticeably faster. In the bottom panel, we compare the evaluation time to the original time-domain \texttt{SEOBNRv4T} waveform model and find it is $\sim 1500$--6000 times faster.

@@

As a final test of the surrogate, we perform several parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template. We use the design sensitivity PSDs for the two advanced LIGO detectors and the advanced Virgo detector~\cite{Aasi:2013wya}. We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the EOS named MS1b in~\cite{ReadLackey2009}. The sky location and inclination angle are fixed at an arbitrary value, and the distances are chosen such that the optimal network signal-to-noise ratios (SNRs) are either 30, 60, or 120. We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.

As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem
\begin{equation}
p(\vec\vartheta | \vec d) \propto p(\vec\vartheta) \mathcal{L}(\vec d | \vec\vartheta),
\end{equation}
where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood. 
For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$. 
As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz. We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately $10^7$ iterations to produce 10,000--20,000 statistically independent samples.

We note that the \texttt{SEOBNRv4T} waveform model still has some signal above 2048~Hz (see Fig.~\ref{fig:hoff}). We therefore also performed runs with a sampling frequency of 8192~Hz and integrated the likelihood function in the interval $[20, 4096]$~Hz. However, this did not have a noticeable effect on the recovered posteriors.
